Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-15 ind_files/01_Introduction.tex
Minor phrasing changes

\section{Outline of the outline of the rest of this thesis}
This section provides a bullet-point outline for those who are really short of time. A more detailed outline of the thesis is provided in the next section.

@@

\begin{itemize}
	\item \textbf{Part \ref{part_theory}}
	develops mathematical formalism for describing finite fluctuating populations from first principles.
		\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized in the next chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations that vary in one or more \emph{discrete} characters. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations. This chapter is biologically original but uses well understood mathematics that has been used in biological contexts before.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians. This chapter is entirely original, both biologically, and (to the best of my knowledge) in its use of certain mathematical structures that extends and generalize previous statistical physics studies of certain ecological models.
		\end{itemize}
	\item \textbf{Part \ref{part_summary}} presents the major results of all the formalism developed up until this point.
		\begin{itemize}
			\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed.
			\item Section \ref{sec_disc_field_eqns} of chapter \ref{chap_unification} discusses the use of the stochastic field equations formulated in chapter \ref{chap_infD_processes}, connects them to some previous studies in biology, and briefly discusses the use of and knowledge about such mathematical objects in applied mathematics fields such as statistical physics. This section is quite technical due to the nature of the subject.
			\item Chapter \ref{sec_disc} provides a non-technical discussion and summary of all the major results, and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). The mathematically averse can skip all other chapters and only read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)} 
		\end{itemize}
\end{itemize}

\section{Outline of the rest of this thesis}

>>>>

\section{A very brief outline of the rest of this thesis}
This section provides a bullet-point outline for those who are really short of time. A more detailed outline of the thesis is provided in the next section.

@@

\begin{itemize}
	\item \textbf{Part \ref{part_theory}}
	develops mathematical formalism for describing finite fluctuating populations from first principles.
		\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized in the next chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations that vary in one or more \emph{discrete} characters. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations. This chapter is biologically original but uses well understood mathematics that has been used in biological contexts before.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians. This chapter is entirely original, both biologically, and (to the best of my knowledge) in its use of certain mathematical notions that generalize some ideas seen in statistical physics.
		\end{itemize}
	\item \textbf{Part \ref{part_summary}} presents the major results of all the formalism developed up until this point. Those who only want the take home messages and are willing to trust my mathematics can skip to this part.
		\begin{itemize}
			\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed.
			\item Section \ref{sec_disc_field_eqns} of chapter \ref{chap_unification} discusses the use of the stochastic field equations formulated in chapter \ref{chap_infD_processes}, connects them to some previous biological models, and briefly discusses the use of and knowledge about such mathematical objects in other areas of applied mathematics such as statistical physics. This section is quite technical due to the nature of the subject.
			\item Chapter \ref{sec_disc} provides a non-technical discussion and summary of all the major results, and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)} 
		\end{itemize}
\end{itemize}

\section{A more expanded outline of the rest of this thesis}

### 2023-03-15 ind_files/06_nontechnical_discussion.tex
Minor phrasing changes

More precisely, the formalism studied in this thesis predicts a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection'. Noise-induced selection is only seen in finite populations and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, behar_fluctuations-induced_2016, constable_demographic_2016,abu_awad_effects_2018,parsons_pathogen_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{constable_demographic_2016,mcleod_social_2019}. Most recently, a study showed that in a very broad class of competition models, noise-induced selection in finite populations can act in opposition to the direction of natural selection predicted in the infinite population limit~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

>>>>

More precisely, the formalism studied in this thesis predicts a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection'. Noise-induced selection is only seen in finite populations and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010,melbinger_evolutionary_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015, behar_fluctuations-induced_2016, constable_demographic_2016,abu_awad_effects_2018,parsons_pathogen_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{constable_demographic_2016,mcleod_social_2019}. Chapter \ref{chap_unification} explains how the  results of these studies can be explained through noise-induced selection. Most recently, a study showed that in a very broad class of competition models, noise-induced selection in finite populations can act in opposition to the direction of natural selection predicted in the infinite population limit~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.
